Left-linear rules in adhesive categories

### 2024-02-14 icalp/fusione.tex
minori
passata su typo etc. soprattutto appendix, senza biblio

\begin{abstract}
%  When two sequences of steps executed by a computational device can
%  be considered equivalent?  This is a relevant question if one
%  considers concurrent and distributed systems, since the independence
%  of two steps means that they could be executed in any order, and
%  potentially in parallel.
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in parallel.
  % 
  We investigate the issue in the context of the double pushout
  approach to rewriting in the general setting of adhesive categories.
  %, a very general setting for rewriting.  
  More precisely, we focus on left-linear rules, which
  not only consume, preserve and generate entities, as linear rules
  do, but may also ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since the canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to well-behaved classes of left-linear rules.
  %
  Moreover, a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

Among rewriting-based formalisms, those manipulating graph-like
structures proved useful in many settings for representing the
dynamics of distributed systems: the graph serves to represent the
entities and their relations within states, while rewriting steps,
modifying the graph, model computation steps that result in changes to
the system state. The DPO (double-pushout) approach~\cite{EhrigPS73}
is nowadays considered a standard one for these structures, thanks to
its locality (rule application has a local effect on a state,
differently from more recent approaches like
SqPO~\cite{CorradiniHHK06} or PBPO$^+$~\cite{OverbeekER23}) and to its
flexibility: the rules allow to specify that, in a rewriting step,
some entities in the current state are removed, some others are needed
for the rewriting step to occur but remain unaltered and some new
entities can be generated. Concerning concurrency, the chosen notion
is switch (also referred to as shift) equivalence, and independence is formally captured by what is
called the Church-Rosser theorem of DPO
rewriting~\cite[Chapter 3, Section~3.4]{CorradiniMREHL97}. %cite[Section~4]{Ehrig78}

@@

Technically, to acquire the expressiveness needed for capturing
fusions, i.e. the merge of some elements in the state, one is required to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, to the best of our knowledge the
phenomena arising when devising a theory of independence and
concurrency for rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories
%(which capture most ``applications'') 
have not been systematically
explored~\cite{Ehrig1976,EHP:BRfTToHLRS,baldan2011adhesivity},
even if an investigation in the setting of quasi-topoi, considering general
non-linear rules yet restricting their applicability,
% of rules and focussing as far as DPO in concerned on quasi-adhesivity, 
has been presented in~\cite{behr2021concurrency,BehrHK23}.

@@

Unfortunately, for left-linear rewriting systems the characterisation
of switch equivalence based on consistent permutations fails,
% : the existence of a consistent permutation is only necessary but
% not sufficient
hence it cannot be used for proving (1) and (2).
%
This leads to asking whether the two properties above hold.
We show that much of the theory can be retained: globality of independence holds in a
weaker form, conceptually linked to the fact that fusions introduce a
form of disjunctive causality, while consistency
of switching holds unchanged. Finally, we prove that a canonical form
for switching sequences can also be recovered.

The paper has the following structure. 
Section~\ref{sec:ade} recalls the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
Section~\ref{sec:equi} presents the key novelties of our proposal,
first introducing the classical notion of independence and 
a novel axiomatic notion of switchability, further showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define well-switching systems,
and prove that all the classical results holding for linear rewriting systems
can be recovered for such a class of left-linear ones.
Finally, Section~\ref{conclusions} wraps up the paper and 
offers some thoughts on possible directions for future work.
The appendixes collect additional data, recalling basic results on
adhesive categories (Appendix~\ref{app:ade}), introducing and 
proving properties for \emph{strong enforcing} rewriting systems, a class of systems where the local Church-Rosser Theorem holds
(Appendix~\ref{app:fill}), and giving the proofs of the main 
results of our work (Appendix~\ref{omitted}).

@@

\noindent
\parbox{10cm}{
  \begin{definition}[Van Kampen property]
    Let $\X$ be a category and consider the diagrams aside.
 %   
%    \hspace{15pt}
    Given  a class of arrows $\mathcal{A}\subseteq \mor(\X)$, we say that the bottom square
    % \emph{has the Van Kampen property relatively to $\mathcal{A}$}, or that it
    is an \emph{$\mathcal{A}$-Van Kampen square} if
    \begin{enumerate}
    \item it is a pushout square;
    \item 	whenever the cube above has pullbacks as back and left faces and the vertical arrows belong to $\mathcal{A}$, then its top face is a pushout 
    if and only if the front and right faces are pullbacks.
    \end{enumerate}
    Pushout squares which enjoy the ``if'' half of item (2) above are called \emph{$\mathcal{A}$-stable}.

@@

\begin{example}
  \label{ex:adhesive}
  Every topos is adhesive~\cite{lack2006toposes}, thus in particular
  $\cat{Set}$ is adhesive. By the closure properties recalled before,
  every presheaf $[\cat{X},\cat{Set}]$ is adhesive and in particular
  the category $\cat{Graph} = [ E \rightrightarrows V, \cat{Set}]$ is
  adhesive where $E \rightrightarrows {V}$ is the two object category
  with two morphisms $s,t \colon{E} \to {V}$. In a similar way,
  various categories of hypergraphs can be shown to be adhesive, 
  such as term graphs and hierarchical graphs~\cite{CastelnovoGM24}. We can also notice that the category $\cat{sGraphs}$ of simple graphs, i.e. graphs without
  parallel edges, is $\reg{(\cat{sGraphs})}$-adhesive~\cite{BehrHK23} but not quasiadhesive.
\end{example}

@@

$\mathcal{M}$-adhesive categories are the right context in which to
perform abstract rewriting using what is called the Double-Pushout
approach (DPO). We will recall the basic definitions and properties
of this approach to abstract rewriting.

\begin{definition}[\!\!\cite{habel2012mathcal,heindel2009category}]
  Let $\X$ be an $\mathcal{M}$-adhesive category. A \emph{left
    $\mathcal{M}$-linear} rule $\rho$ is a pair $\rho = (l,r)$ of
  arrows with $l: K \to L$ and $r: K \to R$ such that $l$ belongs to
  $\mathcal{M}$.  The rule $\rho$ is \emph{$\mathcal{M}$-linear} if
  $r\in \mathcal{M}$ too.
  %
  \full{A rule $\rho$ is said to be
    \emph{consuming} if $l$ is not an isomorphism.}%
  %
  The object $L$ is
  the \emph{left-hand side}, $R$ is the \emph{right-hand side}, and
  $K$ the \emph{interface}. 
  % \emph{glueing object}.   AC: sempre chiamato interface

  A \emph{left-linear rewriting system} is a pair $(\X, \R)$ where
  $\X$ is an $\mathcal{M}$-adhesive category and $\R$ is a set of left
  $\mathcal{M}$-linear rules. The rewriting system $(\X, \R)$ is
  called $\mathcal{M}$-\emph{linear}\todo{P: giusto?}
  %
  \full{\emph{(consuming)}}
  %
  if every rule in $\R$ is
  $\mathcal{M}$-linear.
  \full{(consuming)}

@@

\noindent
\parbox{8cm}{
  \begin{definition}[abstraction equivalence]
    Let  $\X$ be an $\mathcal{M}$-adhesive category and 
    $(\X, \R)$ a left-linear rewriting system. 
    An \emph{abstraction equivalence} between two derivations $\der{D}$
    and $\der{D'}$ with the same length and
    $r(\der{D})=r(\der{D}')$, is a family of isomorphisms
    $\{\phi_X\}_{X\in \{G_i,D_i\}}$ such that the diagram on the right commutes
    for every $i\in [0, \lgh(\der{D})]$. We say that $\der{D}$ are $\der{D}'$ are \emph{abstraction
      equivalent}, and use the symbol $\der{D}\equiv_a \der{D}'$, if there exists an abstraction equivalence between
    them.
  \end{definition}
}
\parbox{4cm}{\vspace{-1.35em}$\xymatrix@C=40pt{G'_i&D'_i \ar[r]^{g'_i}
    \ar@{>->}[l]_{f'_i}&G'_{i+1}\\ L_i \ar[u]^{m'_i} \ar[d]_{m_i}& K_i
    \ar[u]^{k'_i} \ar[d]_{k_i} \ar[r]^{r_i} \ar@{>->}[l]_{l_i}
    &R_i \ar[u]^{h'_i} \ar[d]_{h_i}\\G_i
    \ar@/_.45cm/[uu]_(.35){\phi_{G_i}}|\hole&D_i\ar@{>->}[l]^{f_i}\ar@/_.45cm/[uu]_(.35){\phi_{D_i}}|\hole
    \ar[r]_{g_i}&G_{i+1}\ar@/_.45cm/[uu]_{\phi_{G_{i+1}}}}$}

@@

\begin{restatable}{proposition}{propUnique}
  \label{prop:unique}
  Let $\X$ be a $\mathcal{M}$-adhesive category. Suppose that the two
  direct derivations $\dder{D}$ and $\dder{D}'$ below, with the same
  match and applying the same left-linear rule $\rho$ are given
  \[\xymatrix{L \ar[d]_{m}& K \ar[d]^{k}\ar@{>->}[l]_{l} \ar[r]^{r} &
      R \ar[d]^{h} & L \ar[d]_{m}& K \ar[d]^{k'}\ar@{>->}[l]_{l}
      \ar[r]^{r} & R \ar[d]^{h'}\\G & \ar@{>->}[l]^{f} D \ar[r]_{g}& H
      & G & \ar@{>->}[l]^{f'} D' \ar[r]_{g'}& H'}\]
  Then there is an abstraction equivalence between $\dder{D}$ and
  $\dder{D}'$, whose first component is $\id{G}$.
\end{restatable}

@@

\begin{definition}[Sequential independence]
  \label{de:sequential-independence}
  Let $\X$ be an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear rewriting system. 
  Let also
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be a derivation of length $2$, with
  $\dder{D}_i$ using rule $\rho_i = (l_i,r_i)$.  An \emph{independence
    pair} between $\dder{D}_0$ and $\dder{D}_1$, is a pair of arrows
  $i_0\colon R_0\to D_1$ and $i_1\colon L_1\to D_0$ such that the
  following diagram commutes
  \[\xymatrix@C=15pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0
      \ar@{.>}@/^.35cm/[drrr]|(.3)\hole_(.4){i_0}
      \ar[dr]|(.3)\hole_{h_0} && L_1 \ar@{.>}@/_.35cm/[dlll]^(.4){i_1}
      \ar[dl]|(.3)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar@{>->}[l]_{l_1}
      \ar[rr]^{r_1} && R_1 \ar[d]^{h_1}\\G_0 && \ar@{>->}[ll]^{f_0}
      D_0 \ar[rr]_{g_0}&& G_1 && \ar@{>->}[ll]^{f_1} D_1
      \ar[rr]_{g_1}&& G_2}\] In this case we say that $\dder{D}_0$ and
  $\dder{D}_1$ are
  % \emph{weakly sequentially independent} if an independence pair
  % exists. If such independence pair is unique we will say that
  % $\dder{D}$ and $\dder{D'}$ are
  \emph{sequentially independent} and write $\dder{D}_0 \seqind \dder{D}_1$.
\end{definition}

@@

\begin{definition}[Switch]
  \label{def:switch}
  Let $\X$ be an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear rewriting system. 
  Let also
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be a derivation of length $2$ made
  by two sequentially independent derivations $\dder{D}_0$,
  $\dder{D}_1$, with independence pair $(i_0, i_1)$, as in the diagram
  on the left below. A \emph{switch} of $\der{D}$ along $(i_0,i_1)$ is
  a derivation $\der{E}=\{\dder{E}_i\}_{i=0}^1$, between the same objects and using the same rules
  in reverse order, as in the diagram on the right below, 
  \[
    \def\objectstyle{\scriptstyle}\def\labelstyle{\scriptstyle}
    \xymatrix@C=1.1mm@R=8mm{
      {L_0} \ar[d]_{m_0}
      & & {K_0} \ar@{>->}[ll]_{l_0} \ar[rr]^{r_0} \ar[d]_{k_0}
      & & {R_0} \ar[drr]|(.35)\hole_(.6){h_0}  \ar@/^.22cm/[drrrrrr]|(.3)\hole^(.75){i_0}
      & & & & 
      % 
      {L_1}\ar[dll]|(.35)\hole^(.6){m_1} \ar@/_.22cm/_(.75){i_1}[dllllll]
      & & {K_1} \ar@{>->}[ll]_{l_1} \ar[rr]^{r_1} \ar[d]^{k_1}
      & & {R_1} \ar[d]^{h_1} \\
      % 
      % 
      {G_0}
      & & {D_0} \ar@{>->}[ll]^{f_0} \ar[rrrr]_{g_0}
      & & & & {G_1} & &
      & &  {D_1} \ar@{>->}[llll]^{f_1} \ar[rr]_{g_1}
      & & {G_2}
    }
    % 
    \xymatrix@C=1.1mm@R=8mm{
      {L_1} \ar[d]_{m_0'}
      & & {K_1} \ar@{>->}[ll]_{l_1} \ar[rr]^{r_1} \ar[d]_{k_0'}
      & & {R_1} \ar[drr]|(.35)\hole_(.6){h_0'}  \ar@/^.22cm/@{.>}^(.75){i_0'}|(.3)\hole[drrrrrr]
      & & & & 
      % 
      {L_0}\ar[dll]|(.35)\hole^(.6){m_1'} \ar@/_.22cm/@{.>}_(.75){i_1'}[dllllll] 
      & & {K_0} \ar@{>->}[ll]_{l_0} \ar[rr]^{r_0} \ar[d]^{k_1'}
      & & {R_0} \ar[d]^{h_1'} \\
      % 
      % 
      {G_0}
      & & {D_0'} \ar@{>->}[ll]^{f_0'} \ar[rrrr]_{g_0'}
      & & & & {G'_1} & &
      & &  {D_1'} \ar@{>->}[llll]^{f_1'} \ar[rr]_{g_1'}
      & & {G_2}  }
    %
  \]
  such that there is an independence pair $(i_0', i_1')$ between
  $\dder{E}_0$ and $\dder{E}_1$ and 
  \begin{center}   
    $m_0=f_0' \circ i_1'$
    \qquad $h_1=g_1' \circ i_0'$
    \qquad $m_0'= f_0 \circ i_1$
    \qquad $h_1'= g_{1}\circ i_0$.
  \end{center}

@@

  A derivation $\der{D}$ consisting of three steps $\dder{D}_0$,
  $\dder{D}_1$, and $\dder{D}_2$, each applying the corresponding rule
  $\rho_i$, is depicted in Fig.~\ref{fi:derD}. Note that the numbers in $\rho_1$
  were changed consistently to represent the vertical morphisms.
  Steps $\dder{D}_1$ and $\dder{D}_2$ are clearly sequential
  independent via the independence pair $(R_1 \to D_2, L_2 \to D_1)$
  mapping nodes according to their numbering.
  %
  A switch of $\der{D}$ along such independence pair is the derivation
  $\der{E}$ in Fig.~\ref{fi:derE}. Instead, in $\der{D}$ the first two
  steps applying $\rho_0$ and $\rho_1$ are not sequential independent,
  intuitively because $\rho_1$ uses the node produced by $\rho_0$ for
  attaching a self-loop.

@@

\begin{definition}
  A \emph{permutation} on
  $\interval[0]{n}$ is a bijection
  $\sigma : \interval[0]{n} \to \interval[0]{n}$. We call it a
  \emph{transposition} when there are $i, j \in \interval{n}$,
  $i \neq j$ such that $\sigma(i)=j$, $\sigma(j) = i$ and
  $\sigma(k) = k$ otherwise. Such a transposition is normally denoted as
  $\transp{i}[j]$.

@@

  We say that two derivations $\der{D}, \der{E}:G\Mapsto H$ are \emph{switch
    equivalent}, written $\der{D}\equiv^{sh} \der{E}$, if there is a switching sequence, possibly empty, $\{\der{D}_{k}\}_{k=0}^m$ such that
      \[\der{D}\equiv_a \der{D}_0 \shift{\transp{i_1}} \der{D}_1 \shift{\transp{i_2}} \ldots \shift{\transp{i_m}}
    \der{D}_m\equiv _a \der{E}\]
\end{definition}

@@

\begin{definition}[Strong Enforcing rewriting systems]
  A left-linear rewriting system is \emph{strong enforcing} if
  every independence pair between two direct derivations is strong.
\end{definition}

@@

Still, there are adhesive rewriting systems which are not strong enforcing, as shown by the following example, which is however somehow pathological.

@@

For these reasons we support the idea that a theory of rewriting for
left-linear rules in adhesive categories should be developed for
systems where the uniqueness of the independence pair is always ensured.

>>>>

\begin{abstract}
%  When two sequences of steps executed by a computational device can
%  be considered equivalent?  This is a relevant question if one
%  considers concurrent and distributed systems, since the independence
%  of two steps means that they could be executed in any order, and
%  potentially in parallel.
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in parallel.
  % 
  We investigate the issue in the context of the double pushout
  approach to rewriting in the general setting of adhesive categories.
  %, a very general setting for rewriting.  
  More precisely, we focus on left-linear rules, which
  not only consume, preserve and generate entities, as linear rules
  do, but may also ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since the canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to well-behaved classes of left-linear rules.
  %
  Moreover, a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems, and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

Among rewriting-based formalisms, those manipulating graph-like
structures proved useful in many settings for representing the
dynamics of distributed systems: the graph serves to represent the
entities and their relations within states, while rewriting steps,
modifying the graph, model computation steps that result in changes to
the system state. The DPO (double-pushout) approach~\cite{EhrigPS73}
is nowadays considered a standard one for these structures, thanks to
its locality (rule application has a local effect on a state,
differently from more recent approaches like
SqPO~\cite{CorradiniHHK06} or PBPO$^+$~\cite{OverbeekER23}) and to its
flexibility: the rules allow to specify that, in a rewriting step,
some entities in the current state are removed, some others are needed
for the rewriting step to occur but remain unaltered, and some new
entities can be generated. Concerning concurrency, the chosen notion
is switch (also referred to as shift) equivalence, and independence is formally captured by what is
called the Church-Rosser theorem of DPO
rewriting~\cite[Chapter 3, Section~3.4]{CorradiniMREHL97}. %cite[Section~4]{Ehrig78}

@@

Technically, to acquire the expressiveness needed for capturing
fusions, i.e. the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, %to the best of our knowledge 
the phenomena arising when devising a theory of independence and
concurrency for rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories
%(which capture most ``applications'') 
have not been systematically
explored~\cite{Ehrig1976,EHP:BRfTToHLRS,baldan2011adhesivity},
even if an investigation in the setting of quasi-topoi, considering
non-linear rules yet restricting their applicability,
% of rules and focussing as far as DPO in concerned on quasi-adhesivity, 
has been presented in~\cite{behr2021concurrency,BehrHK23}.

@@

Unfortunately, for left-linear rewriting systems the characterisation
of switch equivalence via consistent permutations fails,
% : the existence of a consistent permutation is only necessary but
% not sufficient
hence it cannot be used for proving (1) and (2).
%
This leads to asking if the two properties above hold.
We show that much of the theory can be retained: globality of independence holds in a
weaker form, conceptually linked to the fact that fusions introduce a
form of disjunctive causality, while consistency
of switching holds unchanged. Finally, we prove that a canonical form
for switching sequences can also be recovered.

The paper has the following structure. 
Section~\ref{sec:ade} recalls the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
Section~\ref{sec:equi} presents the key novelties of our proposal,
first introducing the classical notion of independence and 
a novel axiomatic notion of switchability, further showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define a subclass of left-linear rewriting systems
that we call well-switching rewriting systems,
and prove that all the classical results holding for linear rewriting systems
can be recovered for such a class.
Finally, Section~\ref{conclusions} wraps up the paper and 
offers some thoughts on possible directions for future work.
The appendixes collect additional data, recalling basic results on
adhesive categories (Appendix~\ref{app:ade}), introducing and 
proving properties for \emph{strong enforcing} rewriting systems, a class of systems where the local Church-Rosser Theorem holds
(Appendix~\ref{app:fill}), and giving the proofs of the main 
results of our work (Appendix~\ref{omitted}).

@@

\noindent
\parbox{10cm}{
  \begin{definition}[Van Kampen property]
    Let $\X$ be a category and consider the diagrams aside.
 %   
%    \hspace{15pt}
    Given  a class of arrows $\mathcal{A}\subseteq \mor(\X)$, we say that the bottom square
    % \emph{has the Van Kampen property relatively to $\mathcal{A}$}, or that it
    is an \emph{$\mathcal{A}$-Van Kampen square} if
    \begin{enumerate}
    \item it is a pushout square;
    \item 	whenever the cube above has pullbacks as back and left faces and the vertical arrows belong to $\mathcal{A}$, then its top face is a pushout 
    if and only if the front and right faces are pullbacks.
    \end{enumerate}
    Pushout squares that enjoy only the ``if'' half of item (2) above are called \emph{$\mathcal{A}$-stable}.

@@

\begin{example}
  \label{ex:adhesive}
  Every topos is adhesive~\cite{lack2006toposes}, thus in particular
  $\cat{Set}$ is adhesive. By the closure properties recalled before,
  every presheaf $[\cat{X},\cat{Set}]$ is adhesive and in particular
  the category $\cat{Graph} = [ E \rightrightarrows V, \cat{Set}]$ is
  adhesive where $E \rightrightarrows {V}$ is the two objects category
  with two morphisms $s,t \colon{E} \to {V}$. In a similar way,
  various categories of hypergraphs can be shown to be adhesive, 
  such as term graphs and hierarchical graphs~\cite{CastelnovoGM24}. We can also notice that the category $\cat{sGraphs}$ of simple graphs, i.e. graphs without
  parallel edges, is $\reg{(\cat{sGraphs})}$-adhesive~\cite{BehrHK23} but not quasiadhesive.
\end{example}

@@

$\mathcal{M}$-adhesive categories are the right context in which to
perform abstract rewriting using what is called the Double-Pushout
 (DPO) approach. We will now recall the basic definitions and properties
of this approach. %to abstract rewriting.

\begin{definition}[\!\!\cite{habel2012mathcal,heindel2009category}]
  Let $\X$ be an $\mathcal{M}$-adhesive category. A \emph{left
    $\mathcal{M}$-linear} rule $\rho$ is a pair $\rho = (l,r)$ of
  arrows with $l: K \to L$ and $r: K \to R$ such that $l$ belongs to
  $\mathcal{M}$.  The rule $\rho$ is \emph{$\mathcal{M}$-linear} if
  $r\in \mathcal{M}$ too.
  %
  \full{A rule $\rho$ is said to be
    \emph{consuming} if $l$ is not an isomorphism.}%
  %
  The object $L$ is
  the \emph{left-hand side}, $R$ the \emph{right-hand side}, and
  $K$ the \emph{interface}. 
  % \emph{glueing object}.   AC: sempre chiamato interface

  A \emph{left-linear rewriting system} is a pair $(\X, \R)$ where
  $\X$ is an $\mathcal{M}$-adhesive category and $\R$ a set of left
  $\mathcal{M}$-linear rules. The rewriting system $(\X, \R)$ is
  called $\mathcal{M}$-\emph{linear}\todo{P: giusto?}
  %
  \full{\emph{(consuming)}}
  %
  if every rule in $\R$ is
  $\mathcal{M}$-linear.
  \full{(consuming)}

@@

\noindent
\parbox{8cm}{
  \begin{definition}[abstraction equivalence]
    Let  $\X$ be an $\mathcal{M}$-adhesive category and 
    $(\X, \R)$ a left-linear rewriting system. 
    An \emph{abstraction equivalence} between two derivations $\der{D}$
    and $\der{D'}$ with the same length and
    $r(\der{D})=r(\der{D}')$ is a family of isomorphisms
    $\{\phi_X\}_{X\in \{G_i,D_i\}}$ such that the diagram on the right commutes
    for every $i\in [0, \lgh(\der{D})]$. We say that $\der{D}$ are $\der{D}'$ are \emph{abstraction
      equivalent}, and use the symbol $\der{D}\equiv_a \der{D}'$, if there exists an abstraction equivalence between
    them.
  \end{definition}
}
\parbox{4cm}{\vspace{-1.35em}$\xymatrix@C=40pt{G'_i&D'_i \ar[r]^{g'_i}
    \ar@{>->}[l]_{f'_i}&G'_{i+1}\\ L_i \ar[u]^{m'_i} \ar[d]_{m_i}& K_i
    \ar[u]^{k'_i} \ar[d]_{k_i} \ar[r]^{r_i} \ar@{>->}[l]_{l_i}
    &R_i \ar[u]^{h'_i} \ar[d]_{h_i}\\G_i
    \ar@/_.45cm/[uu]_(.35){\phi_{G_i}}|\hole&D_i\ar@{>->}[l]^{f_i}\ar@/_.45cm/[uu]_(.35){\phi_{D_i}}|\hole
    \ar[r]_{g_i}&G_{i+1}\ar@/_.45cm/[uu]_{\phi_{G_{i+1}}}}$}

@@

\begin{restatable}{proposition}{propUnique}
  \label{prop:unique}
  Let $\X$ be a $\mathcal{M}$-adhesive category. Suppose that the two
  direct derivations $\dder{D}$ and $\dder{D}'$ below, with the same
  match and applying the same left-linear rule $\rho$, are given
  \[\xymatrix{L \ar[d]_{m}& K \ar[d]^{k}\ar@{>->}[l]_{l} \ar[r]^{r} &
      R \ar[d]^{h} & L \ar[d]_{m}& K \ar[d]^{k'}\ar@{>->}[l]_{l}
      \ar[r]^{r} & R \ar[d]^{h'}\\G & \ar@{>->}[l]^{f} D \ar[r]_{g}& H
      & G & \ar@{>->}[l]^{f'} D' \ar[r]_{g'}& H'}\]
  Then there is an abstraction equivalence between $\dder{D}$ and
  $\dder{D}'$, whose first component is $\id{G}$.
\end{restatable}

@@

\begin{definition}[Sequential independence]
  \label{de:sequential-independence}
  Let $\X$ be an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear rewriting system. 
  Let also
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be a derivation of length $2$, with
  $\dder{D}_i$ using rule $\rho_i = (l_i,r_i)$.  An \emph{independence
    pair} between $\dder{D}_0$ and $\dder{D}_1$ is a pair of arrows
  $i_0\colon R_0\to D_1$ and $i_1\colon L_1\to D_0$ such that the
  following diagram commutes
  \[\xymatrix@C=15pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0
      \ar@{.>}@/^.35cm/[drrr]|(.3)\hole_(.4){i_0}
      \ar[dr]|(.3)\hole_{h_0} && L_1 \ar@{.>}@/_.35cm/[dlll]^(.4){i_1}
      \ar[dl]|(.3)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar@{>->}[l]_{l_1}
      \ar[rr]^{r_1} && R_1 \ar[d]^{h_1}\\G_0 && \ar@{>->}[ll]^{f_0}
      D_0 \ar[rr]_{g_0}&& G_1 && \ar@{>->}[ll]^{f_1} D_1
      \ar[rr]_{g_1}&& G_2}\] In this case we say that $\dder{D}_0$ and
  $\dder{D}_1$ are
  % \emph{weakly sequentially independent} if an independence pair
  % exists. If such independence pair is unique we will say that
  % $\dder{D}$ and $\dder{D'}$ are
  \emph{sequentially independent} and write $\dder{D}_0 \seqind \dder{D}_1$.
\end{definition}

@@

\begin{definition}[Switch]
  \label{def:switch}
  Let $\X$ be an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear rewriting system. 
  Let also
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be a derivation made
  by two sequentially independent derivations $\dder{D}_0$,
  $\dder{D}_1$ with independence pair $(i_0, i_1)$, as in the diagram
  on the left below. A \emph{switch} of $\der{D}$ along $(i_0,i_1)$ is
  a derivation $\der{E}=\{\dder{E}_i\}_{i=0}^1$, between the same objects and using the same rules
  in reverse order, as in the diagram on the right below
  \[
    \def\objectstyle{\scriptstyle}\def\labelstyle{\scriptstyle}
    \xymatrix@C=1.1mm@R=8mm{
      {L_0} \ar[d]_{m_0}
      & & {K_0} \ar@{>->}[ll]_{l_0} \ar[rr]^{r_0} \ar[d]_{k_0}
      & & {R_0} \ar[drr]|(.35)\hole_(.6){h_0}  \ar@/^.22cm/[drrrrrr]|(.3)\hole^(.75){i_0}
      & & & & 
      % 
      {L_1}\ar[dll]|(.35)\hole^(.6){m_1} \ar@/_.22cm/_(.75){i_1}[dllllll]
      & & {K_1} \ar@{>->}[ll]_{l_1} \ar[rr]^{r_1} \ar[d]^{k_1}
      & & {R_1} \ar[d]^{h_1} \\
      % 
      % 
      {G_0}
      & & {D_0} \ar@{>->}[ll]^{f_0} \ar[rrrr]_{g_0}
      & & & & {G_1} & &
      & &  {D_1} \ar@{>->}[llll]^{f_1} \ar[rr]_{g_1}
      & & {G_2}
    }
    % 
    \xymatrix@C=1.1mm@R=8mm{
      {L_1} \ar[d]_{m_0'}
      & & {K_1} \ar@{>->}[ll]_{l_1} \ar[rr]^{r_1} \ar[d]_{k_0'}
      & & {R_1} \ar[drr]|(.35)\hole_(.6){h_0'}  \ar@/^.22cm/@{.>}^(.75){i_0'}|(.3)\hole[drrrrrr]
      & & & & 
      % 
      {L_0}\ar[dll]|(.35)\hole^(.6){m_1'} \ar@/_.22cm/@{.>}_(.75){i_1'}[dllllll] 
      & & {K_0} \ar@{>->}[ll]_{l_0} \ar[rr]^{r_0} \ar[d]^{k_1'}
      & & {R_0} \ar[d]^{h_1'} \\
      % 
      % 
      {G_0}
      & & {D_0'} \ar@{>->}[ll]^{f_0'} \ar[rrrr]_{g_0'}
      & & & & {G'_1} & &
      & &  {D_1'} \ar@{>->}[llll]^{f_1'} \ar[rr]_{g_1'}
      & & {G_2}  }
    %
  \]
  such that there is an independence pair $(i_0', i_1')$ between
  $\dder{E}_0$ and $\dder{E}_1$ and 
  \begin{center}   
    $m_0=f_0' \circ i_1'$
    \qquad $h_1=g_1' \circ i_0'$
    \qquad $m_0'= f_0 \circ i_1$
    \qquad $h_1'= g_{1}\circ i_0$.
  \end{center}

@@

  A derivation $\der{D}$ consisting of three steps $\dder{D}_0$,
  $\dder{D}_1$, and $\dder{D}_2$, each applying the corresponding rule
  $\rho_i$, is depicted in Fig.~\ref{fi:derD}. Note that the numbers in $\rho_1$
  were changed consistently to represent the vertical morphisms.
  Steps $\dder{D}_1$ and $\dder{D}_2$ are clearly sequential
  independent via the independence pair $(R_1 \to D_2, L_2 \to D_1)$,
  mapping nodes according to their numbering.
  %
  A switch of $\der{D}$ along such independence pair is the derivation
  $\der{E}$ in Fig.~\ref{fi:derE}. Instead, in $\der{D}$ the first two
  steps applying $\rho_0$ and $\rho_1$ are not sequential independent,
  intuitively because $\rho_1$ uses the node produced by $\rho_0$ for
  attaching a self-loop.

@@

\begin{definition}
  A \emph{permutation} on
  $\interval[0]{n}$ is a bijection
  $\sigma : \interval[0]{n} \to \interval[0]{n}$. We call it a
  \emph{transposition} when there are $i, j \in \interval{n}$,
  $i \neq j$ such that $\sigma(i)=j$, $\sigma(j) = i$, and
  $\sigma(k) = k$ otherwise. Such a transposition is normally denoted as
  $\transp{i}[j]$.

@@

  We say that two derivations $\der{D}, \der{E}:G\Mapsto H$ are \emph{switch
    equivalent}, written $\der{D}\equiv^{sh} \der{E}$, if there exists a switching sequence, possibly empty, $\{\der{D}_{k}\}_{k=0}^m$ such that
      \[\der{D}\equiv_a \der{D}_0 \shift{\transp{i_1}} \der{D}_1 \shift{\transp{i_2}} \ldots \shift{\transp{i_m}}
    \der{D}_m\equiv _a \der{E}\]
\end{definition}

@@

\begin{definition}[Strong enforcing rewriting systems]
  A left-linear rewriting system is \emph{strong enforcing} if
  every independence pair between two direct derivations is strong.
\end{definition}

@@

Still, there are adhesive rewriting systems that are not strong enforcing, as shown by the following example, which is however somehow pathological.

@@

For these reasons we believe a theory of rewriting for
left-linear rules in adhesive categories should be developed for
systems where the uniqueness of the independence pair is ensured.
